Preventing a Nuclear ISIL: \\Recommendations for United States Policy

### 2015-05-10 draft.tex
Added in Policy Section 6.6 - "Theft".

The ISIL nuclear threat is, by definition, a low-probability/high-impact case as this report has shown.  Developing nuclear weapons amidst crippling sanctions, airstrikes, and a dedicated opposition has never been done, despite several nation states attempting to do so.  Therefore, we choose to not repeat the mistakes of history and propose that no changes to U.S. policy be implemented to prevent ISIL from obtaining NWs.

@@

A basic assumption in our policy initiative is the ability of the IC to gather information through the use of radar, traditional and covert human sources, satellites, aircrafts, and ship signals \cite{Richelson2011}. This yields knowledge about the key primary, secondary and tertiary players involved, insight into the ideology and organization of ISIL, entities that might be providing it with economic or military aid, and operational details for the employment of a NW.

@@

The targeted capability of concern for this policy is the theft or diversion of SNM for a nuclear weapon or the theft or diversion of a complete warhead. The pathway intended for intercept is through theft or purchase, and the interception method is IAEA monitoring under the NPT Comprehensive Safeguards Agreements and, where possible, the Additional Protocol (AP). The proposal to strengthen this interception method is to negotiate with states at risk of supporting ISIL willingly or unwillingly to accept NPT AP measures to strengthen safeguards on their SNM and, if existing, any NWs the state possesses. This will help to break the links connecting “Steal” and “Purchase” on the logic tree in Figure XXX with any material needed for NW acquisition that is covered in the NPT safeguards, including “Fusion Device”, “Fission Device”, “RDD”, and “SNM”.

The lead U.S. agency responsible for implementing this policy will be the U.S. Department of State (DOS) supported by the U.S. Defense Threat Reduction Agency (DTRA). Preventing ISIL from acquiring a nuclear weapon, or at the very least SNM, falls under the purview of both the Bureau of Counterterrorism (CT) and the Bureau of International Security and Nonproliferation (ISN) within the DOS. Under this policy strategy, ISN is the principal actor in the DOS as they are the primary U.S. interface with the IAEA on the NPT [A]. DTRA is a Combat Support Agency that supports the Combatant Commanders and Services in responding to any WMD threat. One aspect of this support is managing a research and development portfolio to develop tools and capabilities to detect and respond to WMD threats. DTRA is the agency within the U.S. government that works to verify treaties related to WMD are being carried out by all signatories, and thus will be able to support the DOS ISN in this policy [B].

The AP is, “a legal document granting the IAEA complementary inspection authority to that provided in underlying safeguards agreements,” [C]. The AP’s purpose is to enable the IAEA inspectorate to provide assurances on declared and possible undeclared activities in order to have a more complete picture of a state’s overall nuclear program. The IAEA is granted increased access to information and sites under the AP. Of the 188 UN member states whom are party to the NPT, there are a total of 124 APs (plus Taiwan and Greenland) in force. Considering again Figure 2.2, the states of primary concern that are either not already party to the NPT or have not ratified the Additional Protocol include Algeria, Egypt, Sudan, South Sudan, Saudi Arabia, Yemen, Syria, and Pakistan [2, D]. Of these states, the ones with significant nuclear capabilities that are of most concern are Egypt, Syria, and Pakistan, with Pakistan being of the most concern as they are not party to the NPT and have substantial nuclear capabilities, including an existing NW stockpile [D]. DPRK is a wild card for this policy, but is unavailable through NPT and AP negotiation channels as they are not party to the NPT nor are they considered open to U.S. negotiation in general.

Therefore, this proposed policy is to negotiate with states in proximity and with ties to ISIL, particularly Egypt, Syria, and most importantly Pakistan, to accept additional IAEA NPT safeguards under the AP. Pakistan is an interesting challenge as they are not party to the NPT, but there is precedence in the case of India for strengthening IAEA NPT Safeguards through customized AP agreements. The budget for this policy is inherently small, and the schedule is longer term due to the slower mechanisms of politics.  However, it is important to pursue this policy now in order to convince the states of concern that the U.S. is open to negotiation and will not tolerate willing or unwilling support of ISIL. With a limited budget, the tools for negotiation will be so-called “carrots” and “sticks”. Carrots may include increased cooperation and development of peaceful applications of nuclear technology under the NPT regime, or increased economic trade and development with the cooperating state if nuclear energy is not a priority. Sticks may include economic sanctions, and, in more extreme cases, UN-wide sanctions.
References:
[A]	U.S. Department of State, Bureau of International Security and Nonproliferation. Nuclear Nonproliferation Treaty [Online]. Available: http:// http://www.state.gov/t/isn/npt/
[B]	Defense Threat Reduction Agency. Verifying [Online]. Available: http://http://www.dtra.mil/Missions/Verifying.aspx
[C]	International Atomic Energy Agency. (2014, October 14). IAEA Safeguards Overview: Comprehensive Safeguards Agreements and Additional Protocols [Online]. Available: https://www.iaea.org/
[D]	James Martin Center for Nonproliferation Studies. (2015, April 10). Additional Protocol [Online]. Available: http://www.nti.org

@@

While avoiding unintended consequences or wasted resources are compelling reasons to reconsider the proposed policies, perhaps the most compelling reason is to place the threat and proposed policies into context.  It is perhaps a bit generous to even call ISIL obtaining a nuclear weapon a low probability event as there is zero basis in history for it to be able to do so \cite{pollack2014unthinkable,Bracken2013,Reed2010,Agency2004}. Developing nuclear weapons amidst crippling sanctions, airstrikes, and a dedicated opposition has never been done, despite several nation states attempting to do so \cite{Reed2010,Agency2004}.  In fact, the only nation to develop a NW in the time of war was the U.S., and the U.S. mainland was never seriously threatened. Supposing that ISIL was the proverbial \enquote{Black Swan} case, it is worth noting that, with the exception of the U.S., no state has gone from decision to having a deliverable nuclear weapon in under four years \cite{Reed2010}. Even modern conservative estimates place the timeline at 5-6 years for a competent nation state pursuing an optimal path to a nuclear weapon \cite{Harney2006}. Proponents of the threat of nuclear terrorism will likely point to the possibility that ISIL will circumvent the indigenous development route completely and pursue theft or purchase of key components or a complete device.  Again, there is no history to suggest that a nation state would take that risk given the enormous repercussions it would face \cite{pollack2014unthinkable}.  Additionally, anything short of a complete device would still require significant development to weaponized, perhaps three to four years depending on the amount of assistance ISIL received \cite{Harney2006}.

>>>>

The ISIL nuclear threat is, by definition, a low-probability/high-impact case as this report has shown.  Developing nuclear weapons amidst crippling sanctions, airstrikes, and a dedicated opposition has never been done, despite several nation states attempting to do so.  Therefore, we choose to not repeat the mistakes of history and propose that no changes to U.S. policy be implemented to prevent ISIL from obtaining nuclear weapons.

@@

A basic assumption in our policy initiative is the ability of the IC to gather information through the use of radar, traditional and covert human sources, satellites, aircrafts, and ship signals \cite{Richelson2011}. This yields knowledge about the key primary, secondary and tertiary players involved, insight into the ideology and organization of ISIL, entities that might be providing it with economic or military aid, and operational details for the employment of a nuclear weapon.

@@

The targeted capability of concern for this policy is the theft or diversion of SNM for a nuclear weapon or the theft or diversion of a complete warhead. The pathway intended for intercept is through theft or purchase, and the interception method is IAEA monitoring under the NPT Comprehensive Safeguards Agreements and, where possible, the Additional Protocol (AP). The proposal to strengthen this interception method is to negotiate with states at risk of supporting ISIL willingly or unwillingly to accept NPT AP measures to strengthen safeguards on their SNM and, if existing, any nuclear weapons the state possesses. This will help to break the links connecting \enquote{Steal} and \enquote{Purchase} on the logic tree in \autoref{fig:logic_tree} with any material needed for nuclear weapon acquisition that is covered in the NPT safeguards, including \enquote{Thermonuclear Device}, \enquote{Fission Device}, \enquote{RDD}, and \enquote{SNM}.

The lead U.S. agency responsible for implementing this policy will be the U.S. Department of State (DOS) supported by the U.S. Defense Threat Reduction Agency (DTRA). Preventing ISIL from acquiring a nuclear weapon, or at the very least SNM, falls under the purview of both the Bureau of Counterterrorism (CT) and the Bureau of International Security and Nonproliferation (ISN) within the DOS. Under this policy strategy, ISN is the principal actor in the DOS as they are the primary U.S. interface with the IAEA on the NPT \cite{Circular1970}. DTRA is a Combat Support Agency that supports the Combatant Commanders and Services in responding to any WMD threat. One aspect of this support is managing a research and development portfolio to develop tools and capabilities to detect and respond to WMD threats. DTRA is the agency within the U.S. government that works to verify treaties related to WMD are being carried out by all signatories, and thus will be able to support the DOS ISN in this policy \cite{DefenseThreatReductionAgency}.

The AP is \enquote{a legal document granting the IAEA complementary inspection authority to that provided in underlying safeguards agreements} \cite{InternationalAtomicEnergyAgency1997}. The AP's purpose is to enable the IAEA inspectorate to provide assurances on declared and possible undeclared activities in order to have a more complete picture of a state's overall nuclear program. The IAEA is granted increased access to information and sites under the AP. Of the 188 UN member states whom are party to the NPT, there are a total of 124 APs (plus Taiwan and Greenland) in force. Considering again \autoref{fig:affmap}, the states of primary concern that are either not already party to the NPT or have not ratified the Additional Protocol include Algeria, Egypt, Sudan, South Sudan, Saudi Arabia, Yemen, Syria, and Pakistan \cite{DefenseThreatReductionAgency,InventoryofInternationalNonproliferationOrganizationsandRegimes2013}. Of these states, the ones with significant nuclear capabilities that are of most concern are Egypt, Syria, and Pakistan, with Pakistan being of the most concern as they are not party to the NPT and have substantial nuclear capabilities, including an existing nuclear weapon stockpile \cite{InventoryofInternationalNonproliferationOrganizationsandRegimes2013}. North Korea is a wild card for this policy, but is unavailable through NPT and AP negotiation channels as they are not party to the NPT nor are they considered open to U.S. negotiation in general.

Therefore, this proposed policy is to negotiate with states in proximity and with ties to ISIL, particularly Egypt, Syria, and most importantly Pakistan, to accept additional IAEA NPT safeguards under the AP. Pakistan is an interesting challenge as they are not party to the NPT, but there is precedence in the case of India for strengthening IAEA NPT Safeguards through customized AP agreements. The budget for this policy is inherently small, and the schedule is longer term due to the slower mechanisms of politics.  However, it is important to pursue this policy now in order to convince the states of concern that the U.S. is open to negotiation and will not tolerate willing or unwilling support of ISIL. With a limited budget, the tools for negotiation will be so-called \enquote{carrots} and \enquote{sticks}. Carrots may include increased cooperation and development of peaceful applications of nuclear technology under the NPT regime, or increased economic trade and development with the cooperating state if nuclear energy is not a priority. Sticks may include economic sanctions, and, in more extreme cases, UN-wide sanctions.

@@

While avoiding unintended consequences or wasted resources are compelling reasons to reconsider the proposed policies, perhaps the most compelling reason is to place the threat and proposed policies into context.  It is perhaps a bit generous to even call ISIL obtaining a nuclear weapon a low probability event as there is zero basis in history for it to be able to do so \cite{pollack2014unthinkable,Bracken2013,Reed2010,Agency2004}. Developing nuclear weapons amidst crippling sanctions, airstrikes, and a dedicated opposition has never been done, despite several nation states attempting to do so \cite{Reed2010,Agency2004}.  In fact, the only nation to develop a nuclear weapon in the time of war was the U.S., and the U.S. mainland was never seriously threatened. Supposing that ISIL was the proverbial \enquote{Black Swan} case, it is worth noting that, with the exception of the U.S., no state has gone from decision to having a deliverable nuclear weapon in under four years \cite{Reed2010}. Even modern conservative estimates place the timeline at 5-6 years for a competent nation state pursuing an optimal path to a nuclear weapon \cite{Harney2006}. Proponents of the threat of nuclear terrorism will likely point to the possibility that ISIL will circumvent the indigenous development route completely and pursue theft or purchase of key components or a complete device.  Again, there is no history to suggest that a nation state would take that risk given the enormous repercussions it would face \cite{pollack2014unthinkable}.  Additionally, anything short of a complete device would still require significant development to weaponized, perhaps three to four years depending on the amount of assistance ISIL received \cite{Harney2006}.
